DAND P5 - Quality Red Wine by Kangsan Kim
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(dplyr)
library(reshape2)
```

```{r echo=FALSE, Load_the_Data}
redWine <- read.csv("wineQualityReds.csv")
```

The dataset I've decided to explore in this project is Red Wine data. The
dataset contains information on the different red wine characteristics such as
acidity, sugar, pH, and alcohol%.

# Univariate Plots Section


```{r echo=FALSE, Univariate_Plots}
head(redWine[redWine$alcohol>11, ])
```

This first table is a quick view of a list of wines that have an alcohol 
content of greater than 11%.

```{r echo=FALSE}
summary(redWine)
```

This summary shows some introductory information about each column and the 
range of values they carry.

```{r echo=FALSE}
qplot(x = free.sulfur.dioxide, data = redWine, binwidth = 5, xlim = c(0,60),
      main = "Wine and Free Sulfur Dioxide Content",
      xlab = "Free Sulfur Dioxide",
      ylab = "Number of Wines")
```

This chart explores the free sulfur dioxide content amongst the dataset.
Here, we can see that the vast majority has a free sulfur dioxide content of
less than 20.

```{r echo=FALSE}
qplot(x = total.sulfur.dioxide, data = redWine, binwidth = 5, xlim = c(5,175),
      main = "Wine and Total Sulfur Dioxide Content",
      xlab = "Total Sulfur Dioxide",
      ylab = "Number of Wines")
```

This chart explores the total sulfur dioxide content amongst the dataset.
Here, we can see that the vast majority has a total sulfur dioxide content of
less than 50.

```{r echo=FALSE}
qplot(x = pH, data = redWine, xlim = c(2.7, 4.05), binwidth = 0.05,
      main = "Wine Separated by pH Levels",
      xlab = "pH Levels in Wine",
      ylab = "Number of Wines")
```

This plot shows the number of wines by pH level, and it appears most wines 
are between a pH of 3 and 3.5.

```{r echo=FALSE}
qplot(x = density, data = redWine, xlim = c(0.99, 1), binwidth = 0.001,
      main = "Number of Wines by Density",
      xlab = "Density",
      ylab = "Number of Wines")
```

This plot shows the count of wines per density, ranging from 0.99 to 
1, showing in bins of 0.001.

```{r echo=FALSE}
qplot(x = volatile.acidity, data = redWine, binwidth = 0.01, xlim = c(0,1),
      main = "Wine and Volatile Acidity",
      xlab = "Volatile Acidity",
      ylab = "Number of Wines")
```

This chart explores volatile acidity in the dataset. It appears almost like a
normal distribution!

```{r echo=FALSE}
qplot(x = fixed.acidity, data = redWine, binwidth = 0.5,
      main = "Wine and its Fixed Acidity",
      xlab = "Fixed Acidity in Wine",
      ylab = "Number of Wines")
```

This plot shows the amount of wines for each fixed acidity. We can see that
the majority of the wines seem to be between 7 and 8.

```{r echo=FALSE}
qplot(x = alcohol, data = redWine, xlim = c(8,14), binwidth = 0.5,
      main = "Number of Wines by Alcohol Content [%]",
      xlab = "Alcohol Content (%)",
      ylab = "Number of Wines")
```

This shows the number of wines throughout the different levels of alcohol
content. The overwhelming majority of wines are between 9% and 10%.

```{r echo=FALSE}
qplot(x = quality, data = redWine,
      main = "Wine Quality Rating",
      xlab = "Quality Rating",
      ylab = "Number of Wines")
```

This histogram shows a quick summary of the number of wines in each quality
rating. We can see that there is the majority of wines are rated at a 5 or 6.
I wonder what it takes to receive a rating of 8?

```{r echo=FALSE}
qplot(x = chlorides, data = redWine, xlim = c(0.1,0.15),
      main = "Wine and Chlorides",
      xlab = "Chloride Levels",
      ylab = "Number of Wines")
```

The plot above shows the majority of wines with a chloride level between 0.1
and 0.12.

```{r echo=FALSE}
qplot(x = sulphates, data = redWine, binwidth = 0.01, xlim = c(0.3,1.5),
      main = "Wine and Sulphates",
      xlab = "Sulphate Levels",
      ylab = "Number of Wines")
```

It appears that most of the wine has a sulphate level between 0.4 and 0.8.

```{r echo=FALSE}
by(redWine$residual.sugar, redWine$quality, summary)
```

This is a summary of sugar content between the different quality ratings. 
Here, we can see that the average appears to be about the same within each
quality rating, and does not appear to have a direct effect on the rating
itself.

```{r echo=FALSE}
by(redWine$X, redWine$quality, sum)
```

This is just a count of the number of wines in each quality rating

```{r echo=FALSE}
by(redWine$alcohol, redWine$quality, summary)
```

This shows a numerical summary of the alcohol content sorted by the different
quality levels



# Univariate Analysis

### What is the structure of your dataset?

The structure of my dataset are numerical analysis of the data, along with
bar charts that explore one variable at a time.

### What is/are the main feature(s) of interest in your dataset?

The main interest in my dataset is to explore the relationship between
certain characteristics of the wine in relation to the quality rating.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

The overall numerical analysis can help support my investigation into
the relationship between characteristics of wine and its quality.

### Did you create any new variables from existing variables in the dataset?

For this dataset, I did not create any new variables.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

Through my analysis, I did not notice any unusual distributions. This was a
clean and tidy dataset of wines.


# Bivariate Plots Section


```{r echo=FALSE, Bivariate_Plots}
ggplot(aes(factor(quality), alcohol), data = redWine) +
  geom_jitter(alpha = 0.3) +
  geom_boxplot(alpha = 0.5, color = "blue") +
  stat_summary(fun.y = "mean",
               geom = "point",
               color = "red",
               shape = 8,
               size = 4)
```

This plot explores the different levels of alcohol content in the different 
quality ratings. Here we can see that the median alcohol content in wines
is just above 10%.

```{r echo=FALSE}
ggplot(aes(x = pH, y = density), data = redWine) + 
  geom_point(alpha = 1/5) +
  xlim(2.7,4.02) + 
  coord_trans(y = "sqrt")
```

This plot shows the relationship between pH levels and the density of the
wine. We can see that the density trends downward as pH levels rise.

```{r echo=FALSE}
quality_content <- group_by(redWine, quality)
alcohol_content <- group_by(redWine, alcohol)
```

I created two new datasets using dplyr's group_by() method to group
the data by quality ratings and alcohol content.

```{r echo=FALSE}
ggplot(aes(factor(quality), 
           density), 
       data = redWine) +
  geom_jitter(alpha = 0.3) +
  geom_boxplot(alpha = 0.5, color = "orange") +
  stat_summary(fun.y = "mean",
               geom = "point",
               color = "blue",
               shape = 8,
               size = 4)
```

Here, we can see the average alcohol and pH content for each quality
rating, along with the number of wines in each category.

```{r echo=FALSE}
redWine_by_alc <- summarise(alcohol_content,
          pH_mean = mean(pH),
          sugar_mean = mean(residual.sugar),
          density_mean = mean(density),
          sulphate_mean = mean(sulphates),
          quality_mean = mean(quality),
          n = n())

head(redWine_by_alc)
ggplot(aes(alcohol, sugar_mean), data = redWine_by_alc) + 
  geom_point(size = 1) +
  geom_smooth(method = "lm", se = FALSE, size = 1)
```

The functions above show the average of certain wine characteristics,
sorted by alcohol content. We then analyze one variable, sugar, to see
if there is a relationship between the average sugar content and
alcohol content. As you can see, there doesn't seem to be a direct 
relationship.

```{r echo=FALSE}
ggplot(aes(x = alcohol, y = density), data = redWine) + 
  geom_point(color = "orange", alpha = 1/5)
```

However, there does seem to be a relationship with the average density
of wine, and the alcohol content.

```{r echo=FALSE}
cor.test(redWine$fixed.acidity, redWine$volatile.acidity,
         method = c("pearson"))
```

This was a test to see if there is a correlation between the fixed
acidity and the voliatile acidity using the Pearson method. The Pearson
method states that any value above 0.3 or below -0.3 means the two variables are
significantly correlated. The result for this test is -0.256, which is close!

```{r echo=FALSE}
cor.test(redWine$alcohol, redWine$quality,
         method = c("pearson"))
```

This test helps confirm the previous analysis - the quality and alcohol
content seem to be significantly correlated.



# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

I started off trying to seek correlation between the quality of the wine and
its characteristics. I initally thought that alcohol content may have a
relationship with the quality rating. Also, I wanted to explore if any other
characteristics of wine were directly correlated, or effected, by one another.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

One characteristic I thought was interesting was the overall increase in the
alcohol content in relation to the quality rating. While I did imagine a
higher alcohol content would result in people enjoying the wine more, I
thought that it would have a limit, or not be as strongly correlated as it
showed.

### What was the strongest relationship you found?

The strongest relationship that I discovered through a plot seemed to be the
alcohol content and quality rating, while the strongest relationship I
discovered through the cor.test() function was between fixed and volatile
acidity.


# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}
ggplot(aes(x = alcohol, y = pH, color = factor(quality)),
       data = redWine) +
         geom_point(alpha = 1/5) +
  geom_smooth(method = "lm", se = FALSE, size = 1) +
  scale_color_brewer(type = "seq",
                     guide = guide_legend(title = "Quality"))
```

This plot shows explores the alcohol content in wine compared to its pH
level, sorted out by the quality rating. When considering that a dot means
there are 5 wines that have those characteristics, we can see where the
majority of the wines are.

```{r echo=FALSE}
ggplot(aes(x = alcohol, y = sulphates, color = factor(quality)),
       data = redWine) +
         geom_point(alpha = 1/5)
```

This plot shows that the amount of sulphates seem to be in a consistent range
between 0.5 to 1 as the alcohol content increases. It also shows the lighter 
blues towards the higher alcohol content, which also indicates that it's of 
higher quality.

```{r echo=FALSE}
ggplot(aes(x = pH, y = volatile.acidity/fixed.acidity),
       data = redWine) +
          geom_point(alpha = 1/5, color = "orange")
```

Here, we explore the relationship between acidity and pH. We can see that
There is a higher percentage of volatile acidity as the pH increases (which
means it is less acidic). From here, we can infer that fixed acidity is better
for lower pH levels compared to volatile acidity.


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

I wanted to see if there is any relationship between any of the
characteristics of wine and its quality rating, but initially proposed the
idea that there is no one direct relationship. Other than alcohol content,
my additional analysis proves that it is not just one variable that is a
reliable predictor of wine quality.

### Were there any interesting or surprising interactions between features?

It was interesting to see that volatile acidity and its relationship with
pH levels. When thinking of the word "voliatile", you think of movement
and action, which are some similar characterstics when I think of acidity.
So it was interesting to me to see that it actually made the wine more basic.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
ggplot(aes(x = alcohol, y = pH, color = factor(quality)),
       data = redWine) +
         geom_point(alpha = 1/5) +
  geom_smooth(method = "lm", se = FALSE, size = 1) +
  scale_color_brewer(type = "seq",
                     guide = guide_legend(title = "Quality"))
```

### Description One

I chose this plot that compares alcohol content to its respective pH levels,
separated by the wine's quality rating. While we do not see any direct
relationship, we can see that the majority of wines have an alcohol content
of around 10%, with a pH level of between 3.0 and 3.5.


### Plot Two
```{r echo=FALSE, Plot_Two}

ggplot(aes(factor(quality), alcohol), data = redWine) +
  geom_jitter(alpha = 0.3) +
  geom_boxplot(alpha = 0.5, color = "blue") +
  stat_summary(fun.y = "mean",
               geom = "point",
               color = "red",
               shape = 8,
               size = 4)
```

### Description Two

I chose this plot beacuse the first question I had was "does the alcohol
content have anything to do with its rating?" In my experience, people can
just different wines with a heavy bias on its alcohol content. When I saw that
the median was just above 10%, I understood that it was not judged with as
significantly as I thought, even though most of the wine is rated a 5 or 6.

### Plot Three
```{r echo=FALSE, Plot_Three}

ggplot(aes(x = pH, y = volatile.acidity/fixed.acidity),
       data = redWine) +
          geom_point(alpha = 1/5, color = "orange")
```

### Description Three

I chose this plot because it combines everything I have learned so far to
show the great detail of the amount of acidity in relation to the wine's
pH levels.

------

# Reflection

Some struggles I found in exploring this dataset was trying to find a
meaningful relationship between the variables that I can contribute to
the overall quality of the wine. What did go well was disproving this idea
that a singular factor can cause the quality of the wine to go up or down.
It was surprising to see elements of the wine that you would not think to be
linked together to show a dependent relationship. Moving forward, with
datasets like this, additional work can be done, such as including more
characteristics of the wine. This can allow for a greater search into what
makes a particular wine rate higher than another.